\wde{Dithering}{
    Dithering adds noise to the signal which avoids the signal 
    having zeros (which may be problematic when taking the log).
    \begin{equation*}
        y[t] = x[t] + \epsilon[t] \quad\quad \epsilon[t] \sim \mathcal{N}(0, \sigma^2)
    \end{equation*}
}

\wde{DC Offset}{
    Most processing techniques assume that the signal is centered around 0.
    \begin{equation*}
        y[t] = x[t] - \frac{1}{T} \mu_x \quad\quad \mu_x = \frac{1}{T} \sum_{t=1}^{T} x[t]
    \end{equation*}
}

\wde{Pre-emphasis}{
    Emphasize high frequency components.
    \begin{equation*}
        y[t] = x[t] - \alpha x[t-1]
    \end{equation*}
    where $\alpha$ is a constant (usually 0.97)
}

\wde{Discrete Fourier Transform (DFT)}{
    The dot product of the signal with the sinusoids of the frequency (Fourier basis).
    \begin{equation*}
        X[k] = \sum_{t=0}^{T-1} x[t] e^{i 2 \pi k t / T}
    \end{equation*}
    DFT decomposes the signal into frequency components.
    Commonly $\mathcal{F}$ is used to denote the DFT operator.
}

\wt{Properties of DFT}{
    \begin{itemize}
        \item Linearity: $\mathcal{F}(a_1 x_1 + a_2 x_2) = a_1 \mathcal{F}(x_1) + a_2 \mathcal{F}(x_2)$
        \item Shift: If $y[t] = x[t-m]$, then $Y[k] = e^{i 2 \pi k m / T} X[k]$ % TODO: Prove
    \end{itemize}
}

\wde{Windowing}{
    Given a signal $x[t]$, a window function $w[t]$ (Hamming, Hann, etc.)
    \begin{equation*}
        y[t] = x[t] w[t]
    \end{equation*}
}

\wde{Short-Time Fourier Transform (STFT)}{
    Given a signal $x[t]$, a window function $w[t]$, and a hop size $h$, the STFT is given by:
    \begin{equation*}
        X[k, m] = \sum_{t=0}^{T-1} x[t] w[t-m] e^{-i 2 \pi k t / T}
    \end{equation*}
    where $m$ is the current frame index and $k$ is the frequency index.
    This gives a complex spectrogram of the signal.
}
\wde{Complex Spectrogram}{
Given a frequency bin $X[k] = a + bi$,
\begin{itemize}
    \item Real: $\mathcal{R}(X[k]) = a$
    \item Imaginary: $\Im m(X[k]) = b$
    \item Magnitude: $|X[k]| = \sqrt{a^2 + b^2}$
    \item Phase: $\angle X[k] = \arccos\frac{a}{\sqrt{a^2 + b^2}}$
    \item Energy: $P[k] = |X[k]|^2$
\end{itemize}
}

\wde{Mel Spectrogram}{ % TODO: Check
    After obtaining the spectrogram, $X[k]$, and using the mel filter bank, $H_n[k]$, the mel spectrogram is given by:
    \begin{equation*}
        Y[k, n] = \sum_{k=0}^{K-1} |X[k]| H_n[k]
    \end{equation*}
}

\wde{Mel Frequency Cepstral Coefficients (MFCCs)}{
    A common feature extraction technique for speech recognition 
    which is able to capture speaker characteristics and phonetic content.
    \begin{enumerate}
        \item Extract the mel spectrogram $Y[k, n]$
        \item Apply the Discrete Cosine Transform (DCT) to the mel spectrogram.
        \item Take the first $p$ coefficients (usually 13).
    \end{enumerate}
}